Starcode: sequence clustering based on all-pairs search

### 2014-11-05 doc/document.tex
Updated manuscript.

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. In many
of those applications, sequencing errors need to be corrected.
This is particularly important when using random DNA barcodes to
trace and quantify transcripts or lineages, since spurious barcodes
introduce inexistent entities. The reference population of random
barcodes is unknown and in this case, error correction amounts
to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computationally complexity.

@@

Sequencing technologies never achieve perfect precision. For instance,
the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
consisting of substitutions \citep{pmid18660515, pmid21576222}
and the PacBio platform has a 15\% error rate consisting of insertions
and deletions \citep{pmid19023044}. The throughput of such technologies
has recently created additional needs to develop efficient error
correction algorithms.

@@

Here we describe a sequence clustering algorithm focused on error
correction. The algorithm is called ``starcode'' in reference to 
clusters of random barcodes, which typically have a star shape.
Starcode is based on all-pairs search, \textit{i.e.} all the pairs
of sequences below a given Levenshtein distance are identified during
the graph construction phase. Matching is carried out by lossless
filtration, followed by a trie search.  The novelty of the algorithm
is the poucet strategy, which uses the redundancy of alphabetically
sorted sequences to avoid unnecessary recomputations and gain speed.

In this article we present and benchmark starcode. We show that on
real biological datasets, starcode is faster than existing sequence
clustering software. Even though starcode was designed for error
correction, we show that it can be used for other problems. As an
illustration, we use it to identify enriched motifs in a bacterial
genome and in protein-RNA interaction experiments.

@@

\subsection{Lossless filtration}
When a query has no match, it is advantageous to omit the trie
search. To this end, starcode uses a partition
approach similar to that described by \cite{WuManber1992}. If the
query is partitioned in $\tau+1$ segments, every $\tau$-match
contains at least one of those (assuming that all the segments
have length at least $\tau$). Indeed, there are at most $\tau$
differences between the query and the match, so at least one segment
is unmodified. In the $\tau$-match, the position of the shared
segment is at most $\tau$ nucleotides on the left or on the right
of its position in the query.

These observations are the basis of a filtration method that proceeds
as follows. Every time a sequence is added to the trie, its segments
are added to $\tau+1$ different indexes. More specifically, the
first $\tau$ nucleotides of the sequence are removed, and the rest
of the sequence is partitioned in $\tau+1$ segments of sizes
differing by at most 1 (the longer segments always in 3' for
consistency). The first fragments are added to the first index,
the second fragments to the second index \textit{etc}.

Before the search, segments of the query are looked up in
the indexes, and in case no match is found, the trie search is
omitted because this query has no $\tau$-match. If at least one
segment is found, the trie search is performed. As mentioned above,
segments shared between the query and a $\tau$-match can be shifted
up to $\tau$ nucleotides. For this reason, shifted segments are
looked up in the indexes according to the following scheme ensuring
that no match can be missed.
(Figure \ref{fig:filtration}). The rightmost segment is looked up
in the $\tau$+1-th index, the second rightmost segment and the
contiguous segments shifted by 1 nucleotides are are looked up in the
$\tau$-th index and so on until the first segment and its
contiguous segments shifted by up to $\tau$ nucleotides are
looked up in the first index.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State \textbf{Define:} $\tau$
    \State \textbf{Variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$, $lastseq$, $k$
    \State \textbf{Containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $k \gets$ \textsc{compute} lookup word lengths
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \If{at least one $k$-mer of $seq$ is in the lookup table}
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next sequence
    \State $start \gets$ \textsc{length} of shared prefix
    between $seq$ and $lastseq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$,
    $hits$, $pebbles$)
    \EndFor
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $lastseq \gets seq$
    \EndIf
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{insert} $seq$ $k$-mers into the lookup table
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Parallelization}
Queries are sorted and partitioned in contiguous blocks.
The matching step then proceeds in two phases. In the first, a
distinct trie is built from the sequences of each block according
to the algorithm described above. In the second, all the sequence
blocks are queried against the tries. If the queries are partitioned
in $N$ blocks, the first phase consists of $N$ seek and construct
jobs, wile the second consists of $N(N-1)/2$ query jobs. In each
phase, the jobs show little dependency on each other, so the
matching algorithm can be efficiently parallelized provided $N$
is larger than the number of independent threads.

@@

\subsection{Benchmark conditions}
All the tests were performed on a 16-core dual-processor
Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at
1866~Mhz. Command-line parameters were set equivalently in all
softwares to run in single-core mode allowing up to 3 mismatches for
input sequences of length 50. Execution options used in simulation and
real datasets are summarized in Table \ref{tab:params1} and Table
\ref{tab:params2}, respectively.

\begin{table}[h]
\centering
\caption{Software execution options used in simulation benchmark.}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{l|l}
    \bf{software} & \bf{execution options}\\
    \hline
    starcode & \texttt{starcode -d3}\\
    slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
    cd-hit 4.6.1 & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
    seed & \texttt{SEED --mismatch 3}\\
\end{tabular}
%}
\label{tab:params1}
\end{table}

\begin{table}[h]
\centering
\caption{Software execution options used in real data benchmark.}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{l|l}
    \bf{software} & \bf{execution options}\\
    \hline
    starcode & \texttt{starcode -d3}\\ 
    slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
    cd-hit 4.6.1 & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
    seed & \texttt{SEED --mismatch 3 --shift 3}\\ 
\end{tabular}
%}
\label{tab:params2}
\end{table}

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. Input formats may be raw
sequence, FASTA or FASTQ (in which case starcode ignores the quality).
By default, starcode outputs sequence clusters, it can also be used
to filter out redundant sequences from input files. By default,
clustering is performed under the assumption that divergence occurs
from experimental errors (sequencing errors, PCR mutations
\textit{etc}.), a more general algorithm is also available for other
clustering problems (an example of which is given in section
\ref{section:motifs}).

We show the basic performance and scalability of starcode on a
dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. Only one
parameter is modified whereas the others were kept constant.

Figure \ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic
scale the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (the
naive implementation of all-pairs search). Note that the sequences of
this dataset have no match, see section \ref{section:benchmark} for
an evaluation of the performance on more realistic datasets.
Figure \ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing number
of threads in Figure \ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

@@

\subsection{Benchmark}
\label{section:benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, while the other tools run on single
reads, for this reason all the tools could not be run no the same
dataset.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets. We also observed that slidesort found 5-10\% less
3-matches than starcode on all the datasets. We were surprised by
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons is feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc).

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. For seed
and cd-hit it decreases as the cluster size increases, whereas
for starcode and slidesort it increases. In spite of this trend,
the performance of starcode remains competitive, even for clusters
of 1 million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit. The memory usage of starcode is the highest
for clusters of size 1000, but it decreases and becomes lower than
the memory usage of seed as the size of the cluster increases.

In conclusion, starcode was the only tool to achieve perfect precision
on these datasets. It also maintains a competitive performance in
terms of running time and memory usage.

%% Biological datasets %%
The performance on artificial data is not always in agreement with
the performance on experimental datasets. Typical experiments present
additional difficulties. For instance,
the sizes of the clusters may be uneven and the reads may contain near
constant regions that may degrade the performance of the algorithm.
We benchmarked sequence clustering algorithms on the problem of
clustering TRIP barcodes \citep{pmid23953119}. Briefly, the principle
of TRIP (Thousands of Reporters Integrated in Parallel) is to tag
reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no
reference to match aberrant barcodes against, because the tagging
sequences are unknown.

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Datasets 1 and 2
consist of Illumina single reads; they differ by the total read count
and by the empirical cluster sizes. According to the output of starcode,
the largest clusters of dataset 1 contain approximately 2,000 sequences,
while dataset 2 contains 4 clusters with more than 1 million sequences.
Dataset 3 consist of paired-end reads.

\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    dataset 1 & 16,457,527   & 50  & single      \\
    dataset 2 & 127,675,537  & 50  & single      \\
    dataset 3 & 2,460,226    & 100  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

>>>>

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. In many
of those applications, sequencing errors need to be corrected.
This is particularly important when using random DNA barcodes to
trace and quantify transcripts or lineages, since spurious barcodes
introduce inexistent entities. The reference population of random
barcodes is unknown and in this case, error correction amounts
to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computational complexity.

@@

Sequencing technologies never achieve perfect precision. For instance,
the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
consisting of substitutions \citep{pmid18660515, pmid21576222}
and the PacBio platform has a 15\% error rate consisting of insertions
and deletions \citep{pmid19023044}. The enormous throughput of such technologies
has recently created additional needs to develop efficient error
correction algorithms.

@@

Here we describe a sequence clustering algorithm focused on error
correction. The algorithm is called ``starcode'' in reference to 
clusters of random barcodes, which typically have a star shape.
Starcode is based on all-pairs search, \textit{i.e.} all the pairs
of sequences below a given Levenshtein distance are identified during
the graph construction phase. Matching is carried out by lossless
filtration, followed by an exhaustive search on the branches of a
prefix trie.  The novelty of the algorithm is the poucet strategy,
which uses the redundancy of alphabetically sorted sequences to avoid
unnecessary recomputations and gain speed.

In this article we present and benchmark starcode. We show that on
real biological datasets, starcode is orders of magnitude faster than
existing sequence clustering software. Even though starcode was
designed for error correction, we also show that it can be used for
other problems. As an illustration, we use it to identify enriched
motifs in a bacterial genome and in protein-RNA interaction
experiments.

@@

\subsection{Lossless filtration}
When a query has no match, it is advantageous to omit the trie
search. To this end, starcode uses a partition
approach similar to that described by \cite{WuManber1992}. The query
is initially partitioned in $\tau+1$ segments. Assuming that all the
segments have length at least $\tau$, then every $\tau$-match present
in the database will contain at least a verbatim copy of one of the
query segments. Indeed, there are at most $\tau$ editions between
the query and the match to be distributed in $\tau+1$ regions, so at
least one segment is unmodified. Due to potential insertions and
deletions in the preceding segments, the shared segment may be shifted
up to $\tau$ nucleotides on the left (all insertions) or on the right
(all deletions) from its original position in the query.

These observations are the basis of a filtration method with 100\%
specificity. More precisely, the segments are defined as follows:
the first $\tau$ nucleotides of the sequence are removed, and the rest
of the sequence is partitioned in $\tau+1$ segments of sizes differing
by at most 1 (the longer segments always in 3' for consistency). Every
time a sequence is added to the trie, it is partitioned and its
segments are added to $\tau+1$ different indexes. The first fragments
are added to the first index, the second fragments to the second index,
\textit{etc}. Before the search, the query is partitioned in the same
way and its segments are looked up in the indexes. In case no match is
found, this query has no $\tau$-match in the current database,
therefore the trie search is omitted. Conversely, if at least one
segment is found, the trie search must performed.

As mentioned above, segments shared between the query and a
$\tau$-match may be found shifted up to $\tau$ nucleotides. For this
reason, shifted segments of the query are looked up in the indexes
according to the scheme of Figure \ref{fig:filtration}, which ensures
that no match can be missed: the rightmost segment is looked up in the
$\tau$+1-th index, the second rightmost segment and the contiguous
segments shifted by 1 nucleotides are are looked up in the $\tau$-th
index and so on, until the first segment and its contiguous segments
shifted by up to $\tau$ nucleotides are looked up in the first index.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State \textbf{Define:} $\tau$
    \State \textbf{Variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$, $lastseq$, $k$
    \State \textbf{Containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $k \gets$ \textsc{compute} filter segment lengths
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \If{at least one $k$-mer of $seq$ is in the filter index}
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next sequence
    \State $start \gets$ \textsc{length} of shared prefix
    between $seq$ and $lastseq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$,
    $hits$, $pebbles$)
    \EndFor
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $lastseq \gets seq$
    \EndIf
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{insert} $seq$ $k$-mers into the filter index
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Parallelization}
Queries are sorted and partitioned in contiguous blocks.
The matching step then proceeds in two phases. In the build phase, a
distinct trie is built from the sequences of each block according
to the algorithm described above. In the second, all the sequence
blocks are queried against all the other tries. If the queries are
partitioned in $N$ blocks, the first phase consists of $N$ seek and
construct jobs, wile the second consists of $N(N-1)/2$ query jobs. In
each phase, the jobs show no dependency on each other, so the
matching algorithm can be efficiently parallelized provided $N$ is
larger than the number of independent threads.

@@

\subsection{Benchmark conditions}
All the tests were performed on a 16-core dual-processor
Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at
1866~Mhz. Command-line parameters were set equivalently in all
softwares to run in single-core mode allowing up to 3 mismatches for
input sequences of length 50. Tables \ref{tab:params1} and
\ref{tab:params2} summarize the execution options used in simulation
and real data sets, respectively.

\begin{table}[h]
\centering
\caption{Software execution options used in simulation benchmark.}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{l|l}
    \bf{software} & \bf{command-line options}\\
    \hline
    starcode & \texttt{starcode -d3}\\
    slidesort & \texttt{slidesort\_v2 -d 3 -t E -c DNA}\\
    cd-hit-est & \texttt{cd-hit-est -n 9 -c 0.9 -M 0 -r 0}\\
    seed & \texttt{SEED --mismatch 3}\\
\end{tabular}
%}
\label{tab:params1}
\end{table}

\begin{table}[h]
\centering
\caption{Software execution options used in real data benchmark.}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{l|l}
    \bf{software} & \bf{command-line options}\\
    \hline
    starcode & \texttt{starcode -d3}\\ 
    slidesort & \texttt{slidesort\_v2 -d 3 -u -t E -c DNA}\\
    cd-hit-est & \texttt{cd-hit-est -n 8 -c 0.94 -M 0}\\
    seed & \texttt{SEED --mismatch 3 --shift 3}\\ 
    rainbow & \texttt{rainbow cluster -m 3}
\end{tabular}
%}
\label{tab:params2}
\end{table}

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. File formats compatible
with starcode are: raw sequence, raw sequence with count, FASTA or
FASTQ (in which case starcode ignores the quality). Starcode has two
output formats: cluster and non-redundant. In cluster mode, starcode
returns detailed information of the clustering results, i.e. canonical
sequences, cluster sizes and the complete list of their constituent
sequences. In non-redundant mode, only the canonical sequences are
printed. The second mode is especially useful to filter out redundant
sequences from input files. By default, clustering is performed under
the assumption that divergence occurs from experimental errors
(sequencing errors, PCR mutations \textit{etc}.), a more general
algorithm is also available for other clustering problems (an example
of which is given in section \ref{section:motifs}).

We show the basic performance and scalability of starcode on a
dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. In each
test, only one parameter is modified while the others are kept constant.

Figure \ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic
scale the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (the
naive implementation of all-pairs search). Note that the sequences of
this dataset have no match, see section \ref{section:benchmark} for
an evaluation of the performance on more realistic datasets.
Figure \ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing number
of threads in Figure \ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is due to hardware limitations, i.e. there is insufficient
memory bandwidth to satisfy the increased demand of parallel memory accesses.

@@

\subsection{Benchmark}
\label{section:benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, whereas the other tools run on single
reads, for this reason all the tools could not be run no the same
dataset.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets (Figure \ref{fig:benchmark}a). We also observed that
slidesort found 5-10\% less 3-matches than starcode on all the
datasets (Figure \ref{fig:benchmark}b). We were surprised by 
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons are feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc).

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show a linear and sub-linear increase,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this trend, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Comparing with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering, which is by far the most memory
demanding structure. The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

In conclusion, starcode was the only tool to achieve perfect precision
on these datasets at a price of increased memory
footprint. Considering the exactness of the output, starcode maintains
a competitive performance in terms of running time.

%% Biological datasets %%
The performance on artificial data is not always in agreement with
the performance on experimental datasets. Typical experiments present
additional difficulties. For instance,
the sizes of the clusters may be uneven and the reads may contain near
constant regions that usually degrade the performance of filter-based
algorithms. We benchmarked sequence clustering algorithms on the
problem of clustering TRIP barcodes \citep{pmid23953119}. Briefly, the
principle of TRIP (Thousands of Reporters Integrated in Parallel) is
to tag reporter transcripts with random barcodes and measure the
abundance of barcodes in the RNA as a proxy for gene expression. There
is no reference to match aberrant barcodes against, because the
tagging sequences are unknown.

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Datasets 1 and 2
consist of Illumina single reads; they differ by the total read count
and by the empirical cluster sizes. According to the output of starcode,
the largest clusters of dataset 1 contain approximately 2,000 sequences,
while dataset 2 contains 4 clusters with more than 1 million sequences.
Dataset 3 consists of paired-end reads.

\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    dataset 1 & 16,457,527   & 50  & single      \\
    dataset 2 & 127,675,537  & 50  & single      \\
    dataset 3 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}
